Obsidian:\\A Domain Specific Embedded Language\\for Parallel Programming of Graphics Processors

### 2013-10-06 src2/hl/paperThesis.tex
hacking

Obsidian is an embedded language for GPU programming that aims to raise the 
level of abstraction for the GPU programmer, while maintaining control over 
the low-level details that are needed for performance. 
So far,
we have chosen not to try to apply automatic transformations or compiler 
optimisation techniques to a GPU architecture agnostic program and hope 
to get performance. In Obsidian, we want to put the tools to make the 
correct decisions into the hands of the programmer. From Obsidian programs, 
CUDA kernels are generated, while host code must still be written in CUDA.
For key building blocks like scan and reduce, we see a trend towards
the use of fewer and much more complicated kernels~\cite{merrill}, so
that the host code is often relatively simple. We wish to support the construction of these sophisticated kernels.

@@

Another possible use of Obsidian could be as the code generating backend of 
a domain specific language that could 
benefit from GPU acceleration. Reference ~\cite{BEAUTY} is an example of this.   
%One such experiment can be found in 
%

@@

The GPUs we target with Obsidian are NVIDIA GPUs that support CUDA~\cite{CUDA}.
CUDA is the NVIDIA C-dialect for data-parallel programming on their GPUs. 
We try to stay within the subset of functionality that is common 
between CUDA and OpenCL~\cite{OpenCL}, as we may in future want to generate 
OpenCL programs.

@@

There are two different array representations, {\em Pull} and 
{\em Push} arrays. A pull array naturally represents gather operations,
while a push array captures scattering operations. Reference \cite{Obsidian-Expressive} 
describes the introduction of push arrays into Obsidian.

@@

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\cite{COMPILEEDSL} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\cite{FELDSPAR} 
and Repa ~\cite{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as the pull array is long. Hence, 
the {\tt force} function can be used to express parallelism in Obsidian.

@@

The programs that can be specified in Obsidian
are represented in a {\tt Program} data type.
This data type also represents the lowest level programming interface 
exposed by Obsidian. In reference ~\cite{JOSEFJOELCSORT} we make use of these low-level 
capabilities to implement histograms using atomic operations (not shown here).

@@

The choices above influence the compilation procedure. Implementing a compiler 
for the {\tt Program} data type is not entirely straightforward. Our approach 
is described in further detail in reference ~\cite{BB}; the presentation 
here focuses instead on the type directed compilation of the {\tt ForAll} 
primitives.

@@

The {\tt Identifier} constructor does not correspond to any code executed on the GPU. 
It just helps in reifying the {\tt Program} monad. Now it can 
be seen that the value of type {\tt a} that is being returned is the monadic value of 
the {\tt Program}. This follows the same monad reification technique as described 
in reference~\cite{BB}.

@@

\noindent where $z$ and $c$ are complex numbers. The method to generate the fractal 
presented here is based on a sequential C program from reference 
~\cite{Fractals}. However, the problem is embarrassingly parallel.

@@

Some of the reduction 
kernels will also require that the operation is commutative. 
%A typical reduction is to sum up the elements of an array, here this is used 
%as the running example. In this section we implement a series of reduction 
%kernels. 
To illustrate the kind of low level control that an 
Obsidian programmer has over expressing details of a kernel, we show a series 
of reduction kernels, each with different optimisations applied. Many of the 
optimisations that are applied to the kernels are influenced by a 
tutorial presentation from NVIDIA~\cite{reduction}.

@@

Both to avoid conditionals and to allow
for larger scans per block, we move to two elements per thread.
Each phase of the algorithm is a parallel for loop that is executed 
by half as many threads as there are elements to scan. The body of the loop
performs one operation and one copy, using bit-twiddling to compute indices.
Note the use of two {\em write functions} in sequence.
Similar patterns were used in our implementations of sorting networks~\cite{Obsidian-Expressive}, for similar reasons.

@@

%\subsection{Scans} 
%\cite{OLA} \cite{LargeScan}

@@

There are many languages and libraries for GPU programming. Starting at the 
low-level end of the spectrum we have CUDA ~\cite{CUDA}. CUDA is NVIDIA's 
name for the programming model and extended C language for their GPUs. 
It is 
the capabilities of CUDA that we seek to match with Obsidian,
while giving the programmer the benefits of having Haskell as a meta programming 
language. 
%% a high-level and 
%% functional programming language. 
%% \emph {*** Comment from Mary: I think this is not the right place to talk about
%% these limitations. Is it possible (nonetheless) to list what parts of
%% CUDA are (er) lifted to the Obsidian level? Just saying ``benefits of a high
%% level functional programming language seems too vague. We need to list them.]
%% }

@@

While remaining in the imperative world, but going all the way to the 
other end of the high-level - low-level spectrum, we have the NVIDIA
Thrust Library~\cite{THRUST}.
Thrust offers a programming model where details of GPU architecture are 
completely abstracted away. Here, the programmer expresses algorithms 
using building blocks like: \emph{Sort}, \emph{Scan} and \emph{Reduce}.

Data.Array.Accelerate is a language embedded in Haskell for GPU programming 
~\cite{ACCELERATEDAMP11}. The abstraction level is comparable to that of Thrust. In 
other words, Accelerate hides most GPU details from the programmer. Accelerate 
provides a set of operations (that are parallel and suitable for GPU execution, 
much like in Thrust) implemented as skeletons. Recent work has permitted
the optimisation of Accelerate programs using fusion techniques to decrease
the number of kernel invocations needed (see reference ~\cite{ACCOPT}).
%Accelerate programs are being optimised using using fusion 
%techniques to decrease the number of kernel invocations needed. 
It seems to us that when using Accelerate the programmer has no control over how to 
decompose his computation onto the GPU or how to make use of shared memory 
resources. For many users, remaining entirely within Haskell will be
a big attraction of Accelerate.

Nikola ~\cite{NIKOLA} is another language embedded in Haskell that occupies the 
same place as Accelerate and Thrust on the abstraction level spectrum.

The systems above are all for flat data-parallelism, Bergstrom 
and Reppy are attempting nested data-parallelism by implementing a 
compiler for the NESL language for GPUs ~\cite{NESL}.

@@

The Copperhead \cite{copperhead} system compiles a subset of Python to run 
on GPUs. Much like other languages mentioned here, Copperhead identifies 
usages of certain parallel primitives that can be executed in parallel 
on the GPU (such as reduce, scan and map). But Copperhead also 
allows the expression of nested data-parallelism and is in that 
way different from both Accelerate and Obsidian.

In reference ~\cite{Finance}, Oancea et al. use manual transformations
to study a set of 
compiler optimisations for generating efficient GPU code from high-level 
and functional programs based on {\tt map}, {\tt reduce} and {\tt scan}. 
They tackle performance problems related to GPU programming, such as bad 
memory access patterns and diverging branches.
Obsidian enables easy exploration of decisions related to these issues.
%These are concepts that we try 
%to give the Obsidian programmer language support to reason about.

In the implementation of Obsidian we use a very simple and direct approach 
to monad reification. Related to this work is the approach based on a 
continuation monad by Persson et al~\cite{Generic}. Recent work by Sculthorpe et al. in reference ~\cite{sculthorpe2013constrained} is also related. 
Our approach to monad reification is simpler but less general.

@@

In reference ~\cite{CSORT}, we experiment with atomic operations 
and are forced to write our code in a very imperative style. The addition of mutable 
arrays and a set of operations on them could improve this situation 
while giving up purity. We will explore what the addition of mutable arrays to 
Obsidian would mean as future work.

@@

Obsidian lends itself well to the kind of experimentation with low level 
GPU details that allow for the implementation of efficient kernels. This is 
illustrated in section \ref{sec:redOpt}, where we manage to apply techniques
used in the NVIDIA tutorial ~\cite{reduction}. The case study also show hows we 
can compose kernels and thus reuse prior effort. 
%For example, {\tt red4} 
%to {\tt red7} all make use of {\tt red3} for their parallel phases.

@@

Our monad reification method certainly simplified
the implementation of Obsidian, and showed that the method works
well in practice (see reference ~\cite{BB} for more details of the method).

@@

%We hinted at our simple and direct method of monad reification. This is described 
%in more detail in reference ~\cite{BB}. 
%In Obsidian, we do not currently make 
%explicit use of the compositional nature of this method.

@@

%\bibliographystyle{abbrvnat}
%\bibliography{paper}
% The bibliography should be embedded for final submission.

>>>>

Obsidian is an embedded language for GPU programming that aims to raise the 
level of abstraction for the GPU programmer, while maintaining control over 
the low-level details that are needed for performance. 
So far,
we have chosen not to try to apply automatic transformations or compiler 
optimisation techniques to a GPU architecture agnostic program and hope 
to get performance. In Obsidian, we want to put the tools to make the 
correct decisions into the hands of the programmer. From Obsidian programs, 
CUDA kernels are generated, while host code must still be written in CUDA.
For key building blocks like scan and reduce, we see a trend towards
the use of fewer and much more complicated kernels~\citehl{merrill}, so
that the host code is often relatively simple. We wish to support the construction of these sophisticated kernels.

@@

Another possible use of Obsidian could be as the code generating backend of 
a domain specific language that could 
benefit from GPU acceleration. Reference ~\citehl{BEAUTY} is an example of this.   
%One such experiment can be found in 
%

@@

The GPUs we target with Obsidian are NVIDIA GPUs that support CUDA~\citehl{CUDA}.
CUDA is the NVIDIA C-dialect for data-parallel programming on their GPUs. 
We try to stay within the subset of functionality that is common 
between CUDA and OpenCL~\citehl{OpenCL}, as we may in future want to generate 
OpenCL programs.

@@

There are two different array representations, {\em Pull} and 
{\em Push} arrays. A pull array naturally represents gather operations,
while a push array captures scattering operations. Reference \citehl{Obsidian-Expressive} 
describes the introduction of push arrays into Obsidian.

@@

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\citehl{COMPILEEDSL} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\citehl{FELDSPAR} 
and Repa ~\citehl{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as the pull array is long. Hence, 
the {\tt force} function can be used to express parallelism in Obsidian.

@@

The programs that can be specified in Obsidian
are represented in a {\tt Program} data type.
This data type also represents the lowest level programming interface 
exposed by Obsidian. In reference ~\citehl{JOSEFJOELCSORT} we make use of these low-level 
capabilities to implement histograms using atomic operations (not shown here).

@@

The choices above influence the compilation procedure. Implementing a compiler 
for the {\tt Program} data type is not entirely straightforward. Our approach 
is described in further detail in reference ~\citehl{BB}; the presentation 
here focuses instead on the type directed compilation of the {\tt ForAll} 
primitives.

@@

The {\tt Identifier} constructor does not correspond to any code executed on the GPU. 
It just helps in reifying the {\tt Program} monad. Now it can 
be seen that the value of type {\tt a} that is being returned is the monadic value of 
the {\tt Program}. This follows the same monad reification technique as described 
in reference~\citehl{BB}.

@@

\noindent where $z$ and $c$ are complex numbers. The method to generate the fractal 
presented here is based on a sequential C program from reference 
~\citehl{Fractals}. However, the problem is embarrassingly parallel.

@@

Some of the reduction 
kernels will also require that the operation is commutative. 
%A typical reduction is to sum up the elements of an array, here this is used 
%as the running example. In this section we implement a series of reduction 
%kernels. 
To illustrate the kind of low level control that an 
Obsidian programmer has over expressing details of a kernel, we show a series 
of reduction kernels, each with different optimisations applied. Many of the 
optimisations that are applied to the kernels are influenced by a 
tutorial presentation from NVIDIA~\citehl{reduction}.

@@

Both to avoid conditionals and to allow
for larger scans per block, we move to two elements per thread.
Each phase of the algorithm is a parallel for loop that is executed 
by half as many threads as there are elements to scan. The body of the loop
performs one operation and one copy, using bit-twiddling to compute indices.
Note the use of two {\em write functions} in sequence.
Similar patterns were used in our implementations of sorting networks~\citehl{Obsidian-Expressive}, for similar reasons.

@@

%\subsection{Scans} 
%\citehl{OLA} \citehl{LargeScan}

@@

There are many languages and libraries for GPU programming. Starting at the 
low-level end of the spectrum we have CUDA ~\citehl{CUDA}. CUDA is NVIDIA's 
name for the programming model and extended C language for their GPUs. 
It is 
the capabilities of CUDA that we seek to match with Obsidian,
while giving the programmer the benefits of having Haskell as a meta programming 
language. 
%% a high-level and 
%% functional programming language. 
%% \emph {*** Comment from Mary: I think this is not the right place to talk about
%% these limitations. Is it possible (nonetheless) to list what parts of
%% CUDA are (er) lifted to the Obsidian level? Just saying ``benefits of a high
%% level functional programming language seems too vague. We need to list them.]
%% }

@@

While remaining in the imperative world, but going all the way to the 
other end of the high-level - low-level spectrum, we have the NVIDIA
Thrust Library~\citehl{THRUST}.
Thrust offers a programming model where details of GPU architecture are 
completely abstracted away. Here, the programmer expresses algorithms 
using building blocks like: \emph{Sort}, \emph{Scan} and \emph{Reduce}.

Data.Array.Accelerate is a language embedded in Haskell for GPU programming 
~\citehl{ACCELERATEDAMP11}. The abstraction level is comparable to that of Thrust. In 
other words, Accelerate hides most GPU details from the programmer. Accelerate 
provides a set of operations (that are parallel and suitable for GPU execution, 
much like in Thrust) implemented as skeletons. Recent work has permitted
the optimisation of Accelerate programs using fusion techniques to decrease
the number of kernel invocations needed (see reference ~\citehl{ACCOPT}).
%Accelerate programs are being optimised using using fusion 
%techniques to decrease the number of kernel invocations needed. 
It seems to us that when using Accelerate the programmer has no control over how to 
decompose his computation onto the GPU or how to make use of shared memory 
resources. For many users, remaining entirely within Haskell will be
a big attraction of Accelerate.

Nikola ~\citehl{NIKOLA} is another language embedded in Haskell that occupies the 
same place as Accelerate and Thrust on the abstraction level spectrum.

The systems above are all for flat data-parallelism, Bergstrom 
and Reppy are attempting nested data-parallelism by implementing a 
compiler for the NESL language for GPUs ~\citehl{NESL}.

@@

The Copperhead \citehl{copperhead} system compiles a subset of Python to run 
on GPUs. Much like other languages mentioned here, Copperhead identifies 
usages of certain parallel primitives that can be executed in parallel 
on the GPU (such as reduce, scan and map). But Copperhead also 
allows the expression of nested data-parallelism and is in that 
way different from both Accelerate and Obsidian.

In reference ~\citehl{Finance}, Oancea et al. use manual transformations
to study a set of 
compiler optimisations for generating efficient GPU code from high-level 
and functional programs based on {\tt map}, {\tt reduce} and {\tt scan}. 
They tackle performance problems related to GPU programming, such as bad 
memory access patterns and diverging branches.
Obsidian enables easy exploration of decisions related to these issues.
%These are concepts that we try 
%to give the Obsidian programmer language support to reason about.

In the implementation of Obsidian we use a very simple and direct approach 
to monad reification. Related to this work is the approach based on a 
continuation monad by Persson et al~\citehl{Generic}. Recent work by Sculthorpe et al. in reference ~\citehl{sculthorpe2013constrained} is also related. 
Our approach to monad reification is simpler but less general.

@@

In reference ~\citehl{CSORT}, we experiment with atomic operations 
and are forced to write our code in a very imperative style. The addition of mutable 
arrays and a set of operations on them could improve this situation 
while giving up purity. We will explore what the addition of mutable arrays to 
Obsidian would mean as future work.

@@

Obsidian lends itself well to the kind of experimentation with low level 
GPU details that allow for the implementation of efficient kernels. This is 
illustrated in section \ref{sec:redOpt}, where we manage to apply techniques
used in the NVIDIA tutorial ~\citehl{reduction}. The case study also show hows we 
can compose kernels and thus reuse prior effort. 
%For example, {\tt red4} 
%to {\tt red7} all make use of {\tt red3} for their parallel phases.

@@

Our monad reification method certainly simplified
the implementation of Obsidian, and showed that the method works
well in practice (see reference ~\citehl{BB} for more details of the method).

@@

%We hinted at our simple and direct method of monad reification. This is described 
%in more detail in reference ~\citehl{BB}. 
%In Obsidian, we do not currently make 
%explicit use of the compositional nature of this method.

@@

\bibliographystylehl{alpha}
\bibliographyhl{thesis}

% The bibliography should be embedded for final submission.

### 2013-10-06 src2/ifl/GPUs.tex
hacking

In the development of microprocessors, the addition of new cores is now the
way forward, rather than the improvement of single thread performance.
Graphics processing units (GPUs) have moved from being specialised graphics
engines to being suitable to tackle applications with high computational
demands. For a recent survey of the hardware, programming methods and tools,
and successful applications, the reader is referred to~\cite{GPUComputing}.
Figure~\ref{fig:image}, taken from that paper, and due to NVIDIA, shows the
architecture of a modern GPU from NVIDIA. It contains 16 multiprocessors, 
grouped in pairs that share a texture fetch unit (TF in the figure). The 
texture fetch unit is of little importance when using the GPU for general 
purpose computations. Each multiprocessor has 8 stream processors (marked 
SP in the figure). These stream processors has access to 16kB of shared memory.

See reference~\cite{GPUComputing} for information about
the very similar AMD GPU architecture. We have used the NVIDIA architecture,
but developments are similar at AMD. Intel's Larrabee processor points to a
future in which each individual core is considerably more powerful than in
today's GPUs~\cite{larrabee}.

@@

The question of how to program powerful data-parallel processors is likely
to continue to be an interesting one. Unlike for current multicore machines,
the question here is how to keep a large number of small processors
productively occupied. NVIDIA's solution has been to develop the
architecture and the programming model in parallel. The result is called
CUDA -- an extension of C designed to allow developers to exploit the power
of GPUs. Reference~\cite{GPUCudaLuebke} gives a very brief but illuminating
introduction to CUDA for potential new users. The idea is that the user
writes small blocks of straightforward C code, which should then run in
thousands or millions of threads. We borrow the example from the above
introduction. To add two $N \times N$ matrices on a CPU, using C, one would
write something like

>>>>

In the development of microprocessors, the addition of new cores is now the
way forward, rather than the improvement of single thread performance.
Graphics processing units (GPUs) have moved from being specialised graphics
engines to being suitable to tackle applications with high computational
demands. For a recent survey of the hardware, programming methods and tools,
and successful applications, the reader is referred to~\citeifl{GPUComputing}.
Figure~\ref{fig:image}, taken from that paper, and due to NVIDIA, shows the
architecture of a modern GPU from NVIDIA. It contains 16 multiprocessors, 
grouped in pairs that share a texture fetch unit (TF in the figure). The 
texture fetch unit is of little importance when using the GPU for general 
purpose computations. Each multiprocessor has 8 stream processors (marked 
SP in the figure). These stream processors has access to 16kB of shared memory.

See reference~\citeifl{GPUComputing} for information about
the very similar AMD GPU architecture. We have used the NVIDIA architecture,
but developments are similar at AMD. Intel's Larrabee processor points to a
future in which each individual core is considerably more powerful than in
today's GPUs~\citeifl{larrabee}.

@@

The question of how to program powerful data-parallel processors is likely
to continue to be an interesting one. Unlike for current multicore machines,
the question here is how to keep a large number of small processors
productively occupied. NVIDIA's solution has been to develop the
architecture and the programming model in parallel. The result is called
CUDA -- an extension of C designed to allow developers to exploit the power
of GPUs. Reference~\citeifl{GPUCudaLuebke} gives a very brief but illuminating
introduction to CUDA for potential new users. The idea is that the user
writes small blocks of straightforward C code, which should then run in
thousands or millions of threads. We borrow the example from the above
introduction. To add two $N \times N$ matrices on a CPU, using C, one would
write something like

### 2013-10-06 src2/ifl/combinators.tex
hacking

Connection patterns that capture common ways to connect sub-circuits into larger structures have been central to our research on functional and relational languages for hardware design.
Inspired by Backus' FP language, Sheeran's early work on $\mu$FP made use of
{\em combining forms} with geometric interpretations~\cite{SheeranLFP84}. This approach
to capturing circuit regularity was also influenced by contact
with designers of regular array circuits in industry -- see reference~\cite{SheeranJUCS} for an overview of this and much other work on functional programming and hardware design.

Later work on (our) Ruby considered the use of binary relations, rather than functions in specifying hardware~\cite{LyngbyRuby}. Lava builds upon these ideas, but also gains much in expressiveness and flexibility by being embedded in Haskell~\cite{lavaICFP,ClaessenThesis}.
The user writes what look like circuit descriptions, but are in fact circuit {\em generators}. Commonly used {\em connection patterns} are captured by higher order functions.

@@

Lava supports simulation, formal verification and netlist generation from definitions like this. Circuit descriptions are {\em run} (in fact symbolically evaluated) in order to produce an intermediate representation, which is in turn written out in various formats (for fixed size instances). So this is an example of {\em staged programming}~\cite{Taha03}.

The Sklansky construction is one way to implement parallel prefix, and there are many others, see for instance Hinze's excellent survey~\cite{Hinze}.
Those who develop prefix algorithms suitable for hardware implementation use a standard notation to represent
the resulting networks. Data flows from top to bottom and
the least significant input is at top left. Black dots represent
operators.
For example, Figure~\ref{fig:skl32} shows the
recursive Sklansky construction for 32 inputs. 
%
\begin{figure}%
\begin{center}
\includegraphics[bb=0 0 123 748, angle=270, scale=0.4]{./ifl/skl32.pdf}
\caption{The Sklansky construction for $32$ inputs. 
It recursively computes the parallel prefix for each half of the inputs (corresponding to the use of {\tt two} in the definition) and then combines the last output of the lower (left) half with each of the outputs of the upper (right) half. The dotted box outlines the recursive call on the lower half of the inputs.}\label{fig:skl32}
\end{center}
\end{figure}

In this work, we plan to investigate the use of connection patterns, and more generally an emphasis on {\em structure}, in parallel programming.
We have chosen to target GPUs partly because of available expertise among our colleagues at Chalmers, and partly because reading papers about General Purpose GPU (GPGPU) programming gave us a sense of d{\'e}j\`a vu. Programs are illustrated graphically, and bear a remarkable resemblance to circuit modules that we have
generated in the past using Lava. We see an opportunity here, as there is an extensive literature, going back to the 1960s, about implementing algorithms on silicon that may provide clues
about implementing algorithms on GPUs. This literature does not seem to have yet been scrutinised by the Data Parallel Programming or GPGPU communities.
This is possibly because GPUs are moving closer to simply being data parallel machines, and so work on library functions has taken inspiration from earlier work on Data Parallel Programming, such as Blelloch's NESL~\cite{NESL}. But some of the restrictions from the early data parallel machines no longer hold today; for instance broadcasting a value to many processors was expensive in the past, but is much easier to do on modern GPUs. So a construction like Sklansky, which requires such broadcasting, should now be reconsidered, and indeed we have found it to give good results in our initial experiments (writing directly in CUDA). In general, it makes sense to spread the net beyond the standard data parallel programming literature when looking for inspiration in parallel algorithm design. We plan to explore the use of ``old'' circuit design ideas in programming library functions for GPUs.

>>>>

Connection patterns that capture common ways to connect sub-circuits into larger structures have been central to our research on functional and relational languages for hardware design.
Inspired by Backus' FP language, Sheeran's early work on $\mu$FP made use of
{\em combining forms} with geometric interpretations~\citeifl{SheeranLFP84}. This approach
to capturing circuit regularity was also influenced by contact
with designers of regular array circuits in industry -- see reference~\citeifl{SheeranJUCS} for an overview of this and much other work on functional programming and hardware design.

Later work on (our) Ruby considered the use of binary relations, rather than functions in specifying hardware~\citeifl{LyngbyRuby}. Lava builds upon these ideas, but also gains much in expressiveness and flexibility by being embedded in Haskell~\citeifl{lavaICFP,ClaessenThesis}.
The user writes what look like circuit descriptions, but are in fact circuit {\em generators}. Commonly used {\em connection patterns} are captured by higher order functions.

@@

Lava supports simulation, formal verification and netlist generation from definitions like this. Circuit descriptions are {\em run} (in fact symbolically evaluated) in order to produce an intermediate representation, which is in turn written out in various formats (for fixed size instances). So this is an example of {\em staged programming}~\citeifl{Taha03}.

The Sklansky construction is one way to implement parallel prefix, and there are many others, see for instance Hinze's excellent survey~\citeifl{Hinze}.
Those who develop prefix algorithms suitable for hardware implementation use a standard notation to represent
the resulting networks. Data flows from top to bottom and
the least significant input is at top left. Black dots represent
operators.
For example, Figure~\ref{fig:skl32} shows the
recursive Sklansky construction for 32 inputs. 
%
\begin{figure}%
\begin{center}
\includegraphics[bb=0 0 123 748, angle=270, scale=0.4]{./ifl/skl32.pdf}
\caption{The Sklansky construction for $32$ inputs. 
It recursively computes the parallel prefix for each half of the inputs (corresponding to the use of {\tt two} in the definition) and then combines the last output of the lower (left) half with each of the outputs of the upper (right) half. The dotted box outlines the recursive call on the lower half of the inputs.}\label{fig:skl32}
\end{center}
\end{figure}

In this work, we plan to investigate the use of connection patterns, and more generally an emphasis on {\em structure}, in parallel programming.
We have chosen to target GPUs partly because of available expertise among our colleagues at Chalmers, and partly because reading papers about General Purpose GPU (GPGPU) programming gave us a sense of d{\'e}j\`a vu. Programs are illustrated graphically, and bear a remarkable resemblance to circuit modules that we have
generated in the past using Lava. We see an opportunity here, as there is an extensive literature, going back to the 1960s, about implementing algorithms on silicon that may provide clues
about implementing algorithms on GPUs. This literature does not seem to have yet been scrutinised by the Data Parallel Programming or GPGPU communities.
This is possibly because GPUs are moving closer to simply being data parallel machines, and so work on library functions has taken inspiration from earlier work on Data Parallel Programming, such as Blelloch's NESL~\citeifl{NESL}. But some of the restrictions from the early data parallel machines no longer hold today; for instance broadcasting a value to many processors was expensive in the past, but is much easier to do on modern GPUs. So a construction like Sklansky, which requires such broadcasting, should now be reconsidered, and indeed we have found it to give good results in our initial experiments (writing directly in CUDA). In general, it makes sense to spread the net beyond the standard data parallel programming literature when looking for inspiration in parallel algorithm design. We plan to explore the use of ``old'' circuit design ideas in programming library functions for GPUs.

### 2013-10-06 src2/ifl/conclusion.tex
apa
hacking

\section{Conclusion} \label{sec:conc}

Obsidian provides a good interface for experimenting with algorithms on GPUs. 
The earlier version described in~\cite{JMT} showed that it is possible to 
generate efficient CUDA code from the kind of high level descriptions we are 
interested in. For the kernel level, the work in progress described in this paper enhances the  expressive power of Obsidian, extending the range of algorithms that can be described, as well as the degree of control exercised by the user. 
Future work will concentrate on improving the performance of the resulting 
applications, as well as on support for the kernel coordination level.

>>>>

\subsection{Conclusion} \label{sec:conc}

Obsidian provides a good interface for experimenting with algorithms on GPUs. 
The earlier version described in~\citeifl{JMT} showed that it is possible to 
generate efficient CUDA code from the kind of high level descriptions we are 
interested in. For the kernel level, the work in progress described in this paper enhances the  expressive power of Obsidian, extending the range of algorithms that can be described, as well as the degree of control exercised by the user. 
Future work will concentrate on improving the performance of the resulting 
applications, as well as on support for the kernel coordination level.
